---
title: "PvFTa1 and flowering time in twenty Kano common bean accessions"
subtitle: "A reproducible reanalysis of a screen-house trial"
author: "Raphael Odera Onunkwor"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
    embed-resources: true
    df-print: kable
execute:
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: ".."
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, dpi = 150)
source("R/00_packages.R")
```

## Summary

*PvFTa1* is the common bean homologue of *Arabidopsis* FLOWERING LOCUS T and is
treated in the breeding literature as a candidate locus for early flowering
[@gonzalez2016; @bhakta2017]. Twenty *Phaseolus vulgaris* accessions collected
from markets across ten Local Government Areas of Kano State, Nigeria, were
grown in a screen house and scored for growth and flowering. Five accessions
were amplified for a 300 bp *PvFTa1* fragment.

This report asks one question: **does carrying the *PvFTa1* amplicon track
flowering time across this panel?**

The answer is no. The five genotyped accessions span the panel's entire 18-day
range in days to first flowering — they include both the earliest and the
latest accession in the trial — and all five gave the product. A
presence/absence assay at this locus has no discriminating power in this
germplasm.

## The trial

Twenty accessions, two pots each, four seeds per pot thinned to two plants at
week 2, in a completely randomised design: 80 plants. Growth traits were
recorded weekly from week 2 to week 6; days to first flowering and days to 50%
flowering were scored once per pot. The trial ran at the Department of
Biological Sciences, Bells University of Technology, Ota, from November 2024 to
July 2025.

The analysis below starts from the original recording sheet rather than from
the summary table, so every number can be traced to a row of data.

```{r}
#| label: tidy
#| results: hide
source("R/01_tidy_data.R")
```

```{r}
#| label: load
growth_long <- readr::read_csv(file.path(paths$derived, "growth_long.csv"),
                               show_col_types = FALSE)
week6 <- readr::read_csv(file.path(paths$derived, "week6_wide.csv"),
                         show_col_types = FALSE)
flowering <- readr::read_csv(file.path(paths$derived, "flowering.csv"),
                             show_col_types = FALSE)
```

### What the sheet actually contains

The export holds two stacked blocks of 40 rows. Before modelling anything, it
is worth checking whether the two blocks are on a common scale.

```{r}
#| label: session-check
week6 |>
  select(reading, PH, NL, NB, LL, LB, SD) |>
  pivot_longer(-reading, names_to = "trait", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(trait, reading) |>
  summarise(mean = mean(value), .groups = "drop") |>
  pivot_wider(names_from = reading, values_from = mean,
              names_prefix = "block_") |>
  mutate(ratio = round(block_2 / block_1, 2)) |>
  arrange(desc(ratio))
```

They are not. Block 2 runs about three times higher than block 1 for plant
height and about seven times higher for stem diameter. A factor of seven on a
stem diameter is a change of measurement scale, not of plants — two seedlings
thinned from the same pot cannot differ that way. Every model below therefore
carries `reading` as a fixed effect, and stem-diameter results are not
comparable across blocks.

This is the kind of thing that only surfaces by going back to the raw sheet.

## Flowering phenology

```{r}
#| label: pheno
pheno <- flowering |>
  filter(!is.na(dtff)) |>
  group_by(accession_code, accession, pvfta1) |>
  summarise(dtff = first(dtff), dt50 = first(dt50),
            category = first(category), .groups = "drop") |>
  arrange(dtff)

pheno |> select(accession, dtff, dt50, category, pvfta1)
```

Sixteen of the twenty accessions carry a flowering record in the sheet.
Days to first flowering runs from **`r min(pheno$dtff)`** days (Kumbotso3) to
**`r max(pheno$dtff)`** days (Dawanao12), a spread of
**`r max(pheno$dtff) - min(pheno$dtff)`** days.

### A note on zero variance

Both pots of every accession recorded exactly the same flowering dates.

```{r}
#| label: zero-var
flowering |>
  filter(!is.na(dtff)) |>
  group_by(accession) |>
  summarise(dtff_sd = sd(dtff), dt50_sd = sd(dt50), .groups = "drop") |>
  summarise(`accessions with any within-accession variation` =
              sum(dtff_sd > 0 | dt50_sd > 0, na.rm = TRUE),
            `accessions checked` = n())
```

Within-accession variance is exactly zero. An F ratio for accession would have
a zero denominator, so ANOVA and any mean-separation test built on it are
undefined for these traits. Days to flowering is treated here as one fixed
observed value per accession rather than as a sample with error. This is a
limitation of how the trait was scored, not a property of the biology.

## Does PvFTa1 track flowering time?

```{r}
#| label: coverage
genotyped <- pheno |> filter(pvfta1 == "Amplified")

tibble(
  metric = c("Accessions genotyped", "Gave the 300 bp product",
             "Earliest genotyped (days)", "Latest genotyped (days)",
             "Panel range (days)", "Genotyped range (days)",
             "Share of panel range covered"),
  value = c(nrow(genotyped), nrow(genotyped),
            min(genotyped$dtff), max(genotyped$dtff),
            max(pheno$dtff) - min(pheno$dtff),
            max(genotyped$dtff) - min(genotyped$dtff),
            sprintf("%.0f%%", 100 * (max(genotyped$dtff) - min(genotyped$dtff)) /
                      (max(pheno$dtff) - min(pheno$dtff))))
)
```

```{r}
#| label: fig-flowering
#| fig-cap: "Flowering window per accession. Filled point: days to first flowering. Open point: days to 50% flowering. Accessions amplified for PvFTa1 in red."
#| fig-height: 6
pal <- c("Amplified" = "#B4462F", "Not genotyped" = "#9AA5B1")

pheno |>
  mutate(accession = forcats::fct_reorder(accession, dtff)) |>
  ggplot(aes(x = dtff, y = accession, colour = pvfta1)) +
  geom_segment(aes(xend = dt50, yend = accession), linewidth = 1.4, alpha = .55) +
  geom_point(size = 2.6) +
  geom_point(aes(x = dt50), size = 2.6, shape = 21, fill = "white", stroke = 1.1) +
  scale_colour_manual(values = pal, name = NULL) +
  scale_x_continuous(breaks = seq(28, 52, 4)) +
  labs(x = "Days after sowing", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

The red points are not clustered at the early end. They run the length of the
axis: the earliest accession in the trial and the latest are both *PvFTa1*
positive, as is an intermediate one. Whatever separates a 28-day accession from
a 46-day accession here, it is not presence of this amplicon.

Two readings are consistent with that. Either the locus is effectively fixed
across these landraces, or the variation that matters is sequence-level rather
than presence/absence. Both point to the same next experiment: sequence the
amplicon and look for polymorphism, rather than screening more accessions for
the band.

## Are the accessions distinguishable at all?

If the panel showed strong, clean genotype differences in every other trait,
the null *PvFTa1* result would look more surprising. It does not.

### One-way ANOVA and Duncan's test

```{r}
#| label: anova
traits <- c(PH = "Plant height (cm)", NL = "Number of leaves",
            NB = "Number of branches", LL = "Leaf length (cm)",
            LB = "Leaf breadth (cm)", SD = "Stem diameter")

anova_summary <- purrr::map_dfr(names(traits), function(tr) {
  d <- week6 |>
    select(accession, reading, value = all_of(tr)) |>
    filter(!is.na(value)) |>
    mutate(across(c(accession, reading), factor))
  fit <- aov(value ~ accession + reading, data = d)
  broom::tidy(fit) |>
    filter(term == "accession") |>
    transmute(trait = traits[[tr]], df, F = round(statistic, 2),
              p = signif(p.value, 3))
})

anova_summary
```

Only stem diameter clears `p = 0.05`, and that result is an artefact: removing
the between-block scale difference collapses the residual variance, so the test
is really detecting which accessions were measured in which block. Discounting
it, **no growth trait shows a significant accession effect at week 6.**

### Where the variance actually sits

A mixed model puts a number on that. Measurements are nested — accession, then
pot within accession, then the individual plant — and `reading` cuts across all
three.

```{r}
#| label: mixed
vc <- purrr::map_dfr(names(traits), function(tr) {
  d <- week6 |>
    select(accession, pot, reading, value = all_of(tr)) |>
    filter(!is.na(value)) |>
    mutate(across(c(accession, pot, reading), factor))

  fit <- lme4::lmer(value ~ reading + (1 | accession) + (1 | accession:pot),
                    data = d, REML = TRUE,
                    control = lme4::lmerControl(check.conv.singular = "ignore"))

  v <- as.data.frame(lme4::VarCorr(fit))
  pick <- function(g) { x <- v$vcov[v$grp == g]; if (length(x)) x else 0 }
  a <- pick("accession"); p <- pick("accession:pot"); r <- pick("Residual")

  tibble(Trait = traits[[tr]], n = nrow(d),
         Accession = sprintf("%.1f%%", 100 * a / (a + p + r)),
         Pot       = sprintf("%.1f%%", 100 * p / (a + p + r)),
         Residual  = sprintf("%.1f%%", 100 * r / (a + p + r)))
})

vc
```

Accession identity accounts for under 10% of total variance in every
vegetative trait. The overwhelming share sits in the residual — variation
between individual plants and between recording sessions.

This does **not** show the accessions are identical. It shows this trial cannot
resolve vegetative differences between them: two pots per accession, heavy
missing data, and two recording sessions that differ systematically leave very
little power. The honest conclusion is that flowering time is the only
phenotype in this dataset that is sharply structured, and it happens to be the
one the marker was supposed to explain.

## A mechanistic check: how big a difference would it take?

The result above is negative, and negative results are worth pressing on. If
presence of *PvFTa1* does not explain an 18-day spread, how large a difference
in the gene's activity *would*?

Reduce the floral transition to its minimum. Photoperiod drives FT production,
FT accumulates, and the plant commits to flowering when the accumulated signal
crosses a threshold $\theta$:

$$\frac{dF}{dt} = \alpha - \delta F, \qquad F(0) = 0, \qquad \text{flowering when } F = \theta$$

The trial sat at 6.7&deg;N, where daylength varies by under an hour across the
year, so holding $\alpha$ constant within an accession is defensible here in a
way it would not be at temperate latitude.

```{r}
#| label: ode-solve
ft_ode <- function(t, state, parms) {
  with(as.list(c(state, parms)), list(c(dF = alpha - delta * F)))
}
solve_numeric <- function(alpha, delta, times) {
  as.numeric(deSolve::ode(y = c(F = 0), times = times, func = ft_ode,
                          parms = list(alpha = alpha, delta = delta))[, "F"])
}
# Integrating factor gives the closed form.
solve_analytic <- function(alpha, delta, times) {
  (alpha / delta) * (1 - exp(-delta * times))
}

times <- seq(0, 60, by = 0.5)
max(abs(solve_numeric(0.05, 0.03, times) - solve_analytic(0.05, 0.03, times)))
```

The solver and the closed form agree to about $10^{-6}$, which is the check that
licenses using either one.

### What the data can identify

Setting $F(t^\*) = \theta$ and solving,

$$t^\* = -\frac{1}{\delta}\log\!\left(1 - \frac{\theta\delta}{\alpha}\right)$$

$\alpha$ and $\theta$ enter only through their ratio, so no collection of
flowering dates can separate them. Fixing $\theta = 1$ costs nothing and makes
$\alpha$ a rate in units of the threshold. $\delta$ is not identified either by
one date per accession, so it is scanned rather than estimated.

```{r}
#| label: ode-invert
implied_alpha <- function(t_obs, delta, theta = 1) {
  theta * delta / (1 - exp(-delta * t_obs))
}

delta_grid <- 10^seq(-3, 0, length.out = 60)
rate_scan <- tidyr::expand_grid(delta = delta_grid,
                                pheno |> select(accession, dtff, pvfta1)) |>
  mutate(alpha = implied_alpha(dtff, delta))

fold_range <- rate_scan |>
  group_by(delta) |>
  summarise(fold = max(alpha) / min(alpha), .groups = "drop")

tibble(`largest rate ratio the spread requires` = round(max(fold_range$fold), 3))
```

These per-accession rates are a **reparameterisation of the observed dates, not
an independent estimate**: one date fixes one rate exactly.

### The finding

Across every turnover rate considered, the panel's full 28-to-46 day spread
requires at most a **1.63-fold** difference in FT production rate. A change that
small sits comfortably within ordinary regulatory variation. It does not require
a gene to be present in one accession and absent in another.

So the model agrees with the marker data rather than contradicting it: a
conserved locus with quantitative differences in output is sufficient to
generate everything observed.

```{r}
#| label: fig-ode-rates
#| fig-cap: "Implied FT production rate per accession at delta = 0.05. The PvFTa1-amplified accessions are the panel maximum and minimum."
rate_ref <- rate_scan |> filter(abs(delta - 0.05) == min(abs(delta - 0.05)))

rate_ref |>
  mutate(accession = forcats::fct_reorder(accession, alpha)) |>
  ggplot(aes(alpha, accession, colour = pvfta1)) +
  geom_segment(aes(x = min(alpha), xend = alpha, yend = accession),
               linewidth = .5, colour = "grey85") +
  geom_point(size = 3) +
  scale_colour_manual(values = pal, name = NULL) +
  labs(x = expression("Implied production rate, "*alpha), y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

Asked on the mechanistic scale, the comparison gives the same answer as the raw
phenotype:

```{r}
#| label: ode-test
amp <- rate_ref |> filter(pvfta1 == "Amplified")
oth <- rate_ref |> filter(pvfta1 != "Amplified")
tibble(
  `amplified range` = sprintf("%.4f - %.4f", min(amp$alpha), max(amp$alpha)),
  `panel range`     = sprintf("%.4f - %.4f", min(rate_ref$alpha), max(rate_ref$alpha)),
  `Wilcoxon p`      = round(suppressWarnings(wilcox.test(amp$alpha, oth$alpha))$p.value, 3)
)
```

### A model that was deliberately not fitted

Days to 50% flowering would in principle supply a second threshold crossing and
make $\delta$ estimable. It cannot here:

```{r}
#| label: dt50-ties
pheno |> count(dt50, name = "accessions") |> arrange(desc(accessions))
```

With 62% of the panel sharing a single recorded value, the interval between
first and 50% flowering is close to a linear function of the first date by
construction. A fitted second threshold would be reporting that artefact as
biology, so the model stops here.

### What it says to do next

Flowering date is a sensitive readout of production rate: around the panel
median, a 10% difference in $\alpha$ moves flowering by roughly seven days. That
sensitivity is why a presence/absence assay is the wrong instrument. Telling
these accessions apart requires measurement where the variation actually lives,
at the level of expression or sequence.

## Departures from the original write-up

Rebuilding from the sheet surfaced several discrepancies with the original
analysis, recorded here rather than quietly repaired.

| Issue | Detail |
|---|---|
| Growth-trait significance | The original reports significant variability in height, leaf number and stem diameter at week 6. Recomputed, no trait clears `p = 0.05`. |
| Missing flowering records | Garu4, Round13, Wudil15 and Lahadin17 have empty flowering columns in the sheet, but the original table reports values for all twenty. |
| Genotyped set | Methods list accessions 3a, 19a, 2a, 12a and 10a; the results chapter and the gel list Gani8 in place of Ruwan10. This report follows the gel. The finding holds under either set. |
| Scale change | Stem diameter is recorded on different scales in the two blocks and should not be read across them. |
| Absent plants | Eight rows of block 2 are all zeros. Zero is a valid branch count but not a valid height or leaf dimension, so those become missing. |

## Conclusion

Flowering time in this panel spans 18 days and is the one trait the trial
resolves cleanly. Every accession tested for *PvFTa1* carried the 300 bp
fragment, including both phenotypic extremes, so the locus is more conserved
across this germplasm than the phenotypic spread would predict. Presence of the
amplicon is not a usable marker for flowering time here.

The productive next step is sequence-level: amplify and sequence *PvFTa1*
across the full panel and test for polymorphism against flowering date, rather
than scoring further accessions for presence of the band.

## Reproducing this report

```
Rscript R/00_packages.R              # once
quarto render report/pvfta1-phenology.qmd
```

Session information, so the numbers above can be tied to a toolchain:

```{r}
#| label: session
#| code-fold: true
sessionInfo()
```

## References

::: {#refs}
:::
